---
title: "Example code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "05/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
```

# Exercise

## Empty droplets

- Import the raw matrix into R; use `DropletUtils`.
  Make a `SingleCellExperiment` object called `sce`.

```{r}
library(DropletUtils)
sce <- DropletUtils::read10xCounts(samples = c("pbmc5k" = "/project/obds/shared/resources/4_r_single_cell/singlecell_droplets/raw_feature_bc_matrix/"),
  col.names = TRUE
)
  
  

sce
colData(sce)
rowData(sce)
metadata(sce)
sce@assays

```

- Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}
scecounts(sce)
#same thing as 
assay(sce, "counts")


sce <- sce[, colSums(assay(sce, "counts")) > 0]
sce
colData(sce)
rowData(sce)
```

- Run `DropletUtils::emptyDrops()`.

```{r}
set.seed(1)
out <- DropletUtils::emptyDrops(sce,
                                assay.type = "counts",
                                lower = 100, 
                                niters = 10000)
# FDR = false discovery rate - multiple testing correction (p values * by the times you did the calulation, type 1 error)
#p value is likely hood of being an empty droplet

out_df <- as.data.frame(out)

out %>% as.data.frame() %>% 
  filter(Limited == TRUE)
```

**How do you read the output? Which droplets would you consider empty?**

> Answer:
> 
>
>
> 

- Remove empty droplets from the data sets.
  Call the new object `sce_filtered_emptydrops`.

```{r}
positions <- which(out$FDR < 0.01)
#which is giving indexes- we now have positions of the FDR values less than 0.01  
sce_filtered_emptydrops <- sce[, positions]
sce_filtered_emptydrops

remove(positions)
```

## Bonus point

- Draw plots comparing empty droplets and other droplets
  (e.g., library size, mitochondrial content, PCA).

```{r}
plot_data <- tibble(
  library_size = colSums(assay(sce, "counts")),
  droplet_status = out$FDR < 0.01)

  

ggplot(plot_data, aes(droplet_status, library_size, fill = droplet_status)) +
  geom_violin() +
    scale_y_log10()
  
  
  
```

```{r}
is.mt <- 
# table(is.mt)
plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}

sce <- 
sce <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

- Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- 
  
  


all_barcodes <- colnames()

compare_table <- tibble(
  emptydrops = all_barcodes %in%    ,
  cellranger = all_barcodes %in%    ,
)
table(compare_table)
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets.

- Run `scDblFinder::scDblFinder()`.

```{r}
set.seed(1)
sce_scdblfinder <- scDblFinder::scDblFinder(sce_filtered_emptydrops)
colData(sce_scdblfinder)
table(sce_scdblfinder$scDblFinder.class)
```

** How do you read the output ? Which droplets would you consider doublets ?
**

> Answer:sce_scdblfinder$scDblFinder.class shows cells that are classified as doublet or singlet
>
>
>
>

-Remove doublets from the data set.
Call the new object `sce_filtered_scdblfinder`.

```{r}
is.singlet <- sce_scdblfinder$scDblFinder.class == "singlet"
sce_filtered_scdblfinder <- sce_scdblfinder[, is.singlet]
sce_filtered_scdblfinder
```

## Bonus point-Draw plots comparing doublets and other droplets.
(e.g., library size, mitochondrial content, PCA)

```{r}
plot_data <- tibble(
barcodes = colnames(sce_filtered_scdblfinder),
library_size = colSums(assay(sce_filtered_scdblfinder, "counts")),
droplet_status = out$FDR < 0.01
)



ggplot(plot_data,
aes(droplet_status, library_size, fill = droplet_status)) +
geom_violin() +
scale_y_log10()



ggplot() +



```

```{r}
is.mt <-

plot_data <- tibble()
ggplot() +



```

```{r}
#set seed
set.seed(1)
#normalise
sce_scdblfinder <- logNormCounts(sce_scdblfinder)
#run PCA
sce_scdblfinder <- runPCA(sce_scdblfinder)
plot_data <- tibble(
barcode = colnames(sce_scdblfinder),
PC1 =  reducedDim(sce_scdblfinder, "PCA")[, "PC1"],
PC2 = reducedDim(sce_scdblfinder, "PCA")[, "PC2"],
droplet_type = sce_scdblfinder$scDblFinder.class

)

ggplot(plot_data) +
geom_point(aes(
x = PC1,
y = PC2,
colour = droplet_type == "doublet",
colour = "red"
)) +
geom_point(aes(
x = PC1,
y = PC2,
colour = droplet_type == "singlet",
colour = "black"
)) +
theme_cowplot()


```


```{r}
#set seed
set.seed(1)
#normalise
sce_scdblfinder <- logNormCounts(sce_scdblfinder)
#run UMAPA
sce_scdblfinder <-
runUMAP(sce_scdblfinder, dimred = "UMAPA", n_dimred = 20)
plot_data <- tibble(
barcode = colnames(sce_scdblfinder),
UMAP1 =  reducedDim(sce_scdblfinder, "UMAP")[, 1],
UMAP2 = reducedDim(sce_scdblfinder, "UMAP")[, 2],
droplet_type = sce_scdblfinder$scDblFinder.class

)

ggplot(plot_data) +
geom_point(aes(
x = UMAP1,
y = UMAP2,
colour = droplet_type == "doublet",
colour = "red"
)) +
geom_point(aes(
x = UMAP1,
y = UMAP2,
colour = droplet_type == "singlet",
colour = "black"
)) +
theme_cowplot()
  
  
```

- Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

  + You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 

sweep.res.list <- 

sweep.stats <- 
bcmvn <- 
```

```{r}
ggplot() +
  
  
  
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- 
nExp_poi <- 
nExp_poi.adj <- 
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- 
```

```{r}
table(   )
```

```{r, message=FALSE}
seu <- 
```

```{r}
table(   )
```

```{r}
doublets_table <- data.frame(
  
  
  
)
table(doublets_table)
```

# Exercise

## Ambient RNA

- Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}

set.seed(1)
sce_decontX <- celda::decontX(sce_scdblfinder)
colData(sce_decontX)
```

- Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

```{r}

umap <-
  plotDimReduceCluster(
    sce_decontX$decontX_clusters,
    # this is an integer vector of cell cluster lables
    dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
    dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2]
  )
```

- Visualize the percentage of contamination estimated by `decontX` in each cell.
  Use the `plotDecontXContamination()` function.

```{r}
plotDecontXContamination(sce_decontX)
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:
> 
> 

- Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

```{r}

rowData(sce_decontX)
rownames(sce_decontX) <-
  uniquifyFeatureNames(rowData(sce_decontX)$ID, rowData(sce_decontX)$Symbol)

plot_cont <- plotDimReduceFeature(
  decontXcounts(sce_decontX), # logcounts will show the contaminated data
  features = c(
    "CD3D",
    "CD3E",
    "GNLY",
    "LYZ",
    "S100A8",
    "S100A9",
    "CD79A",
    "CD79B",
    "MS4A1",
    "PPBP"
  ),
  dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
  dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2]
  
)

plot_uncont <- plotDimReduceFeature(
  logcounts(sce_decontX), # logcounts will show the contaminated data
  features = c(
    "CD3D",
    "CD3E",
    "GNLY",
    "LYZ",
    "S100A8",
    "S100A9",
    "CD79A",
    "CD79B",
    "MS4A1",
    "PPBP"
  ),
  dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
  dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2]
  
)
```

- Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->
```{r}
# 10 clusters
markers <- list(
  Tcells = c("CD3D", "CD3E"),
  monocytes = c("LYZ", "S100A8", "S100A9"),
  Bcells = c("CD79A", "CD79B", "MS4A1"),
  NKcells = "GNLY",
  megakaryocyte = "PPBP"
  
)
cellTypeMappings <- list(
  Tcells = 1,
  monocytes = 2,
  Bcells = 3,
  NKcells = 5,
  megakaryocyte = 7
  
)

plotDecontXMarkerPercentage(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = c("counts", "decontXcounts")
  
  
  
  
)
```

```{r}
plotDecontXMarkerPercentage(
  sce_decontX,
  
  
  
)
```

```{r}
plotDecontXMarkerPercentage(
  
  
  
  
)
```

- Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r, fig.width=10, fig.height=10}
plotDecontXMarkerExpression(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = c("logcounts", "log_decontXcounts")
)
sce_decontX <-
  logNormCounts(sce_decontX, assay.type = "decontXcounts", name = "log_decontXcounts")
sce_decontX <- logNormCounts(sce_decontX)
assays(sce_decontX)
```
