---
title: "Example code for ggplot2"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

What does `ggplot(diamonds)` do?

```{r}
data('diamonds')
ggplot(diamonds)
```

> Answer:

## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer:

## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + geom_point(alpha = 0.2, size = 0.5)
  
```

- Color data points by `cut`.
  (Copy-paste and extend the code chunk above.)

```{r}
head(diamonds)

  ggplot(diamonds, aes(x = carat, y = price)) + 
    geom_point(alpha = 0.2, 
               size = 0.5,
               aes(colour = cut))
          
```

- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}
#This produces a smooth trend line for each category of cut. 
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point(alpha = 0.2,
             size = 0.5
             ) +
  geom_smooth()


#this gives one smooth mean trend line
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.2,
             size = 0.5,
             aes(colour = cut)) +
  geom_smooth()

```

- Assign that last plot to an object called `obds_diamonds`.

```{r}
obds_diamonds <-
  ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point(alpha = 0.2,
             size = 0.5) +
  geom_smooth()


```


# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer:

# Exercise

## Trend lines

Using the `ChickWeight` data set:

- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
data('ChickWeight')
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Chick)) +
  geom_point() 

```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() 

```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth()


```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer: Diet 3

- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth(method = 'lm') + #There are different smoothing methods that can be found in r, under 'method' argument.
  facet_wrap( ~ Diet, nrow = 2)




```

- Assign that last plot to an object called `obds_chickweight`.

```{r}

obds_chickweight <-
  ggplot(ChickWeight, aes(x = Time, y = weight,    colour = Diet)) +
  geom_point() +
  geom_smooth() +
  facet_wrap( ~ Diet, nrow = 2)

```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}
ggplot2::msleep

```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x = order)) +
  geom_bar()

```

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}


ggplot(msleep, aes(x = order)) +
  geom_bar() +
  theme(axis.text.x = element_text(
    size = 2,
    angle = 45,
    hjust = 1
  ))



```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x = order)) +
  geom_bar() +
  labs(y = "Number of Animals", x = "Order") +
  theme(axis.title = element_text(size = 16),
        axis.text.x = element_text(
          size = 2,
          angle = 45,
          hjust = 1
        ))





```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order, fill = genus)) +
  geom_bar() +
  labs(y = "Number of Animals", x = "Order") +
  theme(
    axis.title = element_text(size = 12),
    axis.text.x = element_text(
      size = 6,
      angle = 45,
      hjust = 1
    ),
    legend.position = 'none'
  )

```

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x = order, fill = genus)) +
  geom_bar() +
  labs(y = "Number of Animals", x = "Order") +
  theme(
    axis.title = element_text(size = 12),
    axis.text.x = element_text(
      size = 6,
      angle = 45,
      hjust = 1
    ),
    legend.key.size = unit(2, 'mm')
  )







```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x = order, fill = genus)) +
  geom_bar() +
  labs(y = "Number of Animals", x = "Order") +
  theme(
    axis.title = element_text(size = 12),
    axis.text.x = element_text(
      size = 6,
      angle = 45,
      hjust = 1
    ),
    legend.key.size = unit(2, 'mm')
  ) +
  guides(fill = guide_legend(ncol = 3))








```

- Add a contour of thin black lines to the bars.

```{r}
ggplot(msleep, aes(x = order, fill = genus)) +
  geom_bar(colour = 'black', linewidth = 0.25) +
  labs(y = "Number of Animals", x = "Order") +
  theme(
    axis.title = element_text(size = 12),
    axis.text.x = element_text(
      size = 6,
      angle = 45,
      hjust = 1
    ),
    legend.key.size = unit(2, 'mm')
  ) +
  guides(fill = guide_legend(ncol = 3))








```

- Assign that last plot to an object called `obds_msleep`.

```{r}
obds_msleep <- ggplot(msleep, aes(x = order, fill = genus)) +
  geom_bar(colour = 'black', linewidth = 0.25) +
  labs(y = "Number of Animals", x = "Order") +
  theme(
    axis.title = element_text(size = 12),
    axis.text.x = element_text(
      size = 6,
      angle = 45,
      hjust = 1
    ),
    legend.key.size = unit(2, 'mm')
  ) +
  guides(fill = guide_legend(ncol = 3))










```

# Exercise

## Plotting grid

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r, fig.height=12, fig.width=12}
library(cowplot)


super_plot <- cowplot::plot_grid(obds_diamonds, obds_chickweight, obds_msleep, labels = c("A", "B", "C"), ncol = 1)

super_plot
```

Using `patchwork`.

```{r, fig.height=12, fig.width=12}
library(patchwork)
obds_diamonds / obds_msleep / obds_chickweight
```

Using `gridExtra`.

```{r, fig.height=12, fig.width=12}
library(gridExtra)
?gridExtra::grid.arrange()

grid.arrange(obds_diamonds, obds_msleep, obds_chickweight)


```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
ggsave("gridArragePlots.png", width = 12, height =12)
```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

```{r}
head(mpg)
?mpg
unique(mpg$manufacturer)
unique(mpg$model)
unique(mpg$displ)
unique(mpg$year)
unique(mpg$cyl)
unique(mpg$trans)
unique(mpg$drv)
unique(mpg$cty)
unique(mpg$hwy)
unique(mpg$fl)
unique(mpg$class)



ggplot(mpg, aes(y = hwy, x = cty)) + 
  geom_point(aes(colour = year), size = 1) +
  facet_wrap(~manufacturer) +
  geom_smooth(method = 'lm')
```
